Syntax and Semantics of Cedille

### 2023-03-16 docs/semantics/paper.tex
docs: semantics: strong normalization for types (sans reduction of terms)
Signed-off-by: Chris Jenkins <cwjnkins@uiowa.edu>

\newcommand{\choice}[0]{\zeta}
\newcommand{\elcap}[0]{\cap}
\newcommand{\abs}[4]{{#1}\, #2\! : \! #3.\, #4}
\newcommand{\absu}[3]{{#1}\, #2.\, #3}
\mathchardef\mhyph="2D % Define a "math hyphen"
\newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
\newcommand{\leadstoc}[0]{\ensuremath{\leadsto_{\mathbf{n}}}}
\newcommand{\leadstocs}[0]{\ensuremath{\leadsto_{\mathbf{n}}^*}}
\newcommand{\nleadstoc}[0]{\ensuremath{\nleadsto_{\mathbf{n}}}}
\newcommand{\tpcheck}[0]{\Leftarrow}
\newcommand{\tpsynth}[0]{\Rightarrow}
\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadstocs}{\Rightarrow}}}
\newcommand{\cbe}[0]{c\beta\eta}
\newcommand{\utp}[0]{\mathcal{U}}
\newcommand{\startcase}[1]{\vspace{#1} \noindent\textbf{\underline{Case:}}}
\newcommand{\startsubcase}[1]{\vspace{#1} \noindent\textbf{\underline{Subcase:}}}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}

@@

\author{Aaron Stump, Chris Jenkins \\
  Computer Science \\
  The University of Iowa \\
\texttt{aaron-stump@uiowa.edu}}

@@

The type theory of Cedille is called the Calculus of Dependent Lambda
Eliminations (CDLE).  This document presents the version of CDLE as of
April 13, 2021.  We have made many changes from the first paper on CDLE~\cite{stump17},
mostly in the form of dropping constructs we discovered (to our
surprise) could be derived~\cite{stump18}. We have also omitted
\emph{lifting} -- a technique for large eliminations with lambda
encodings -- in this document's version of CDLE.  Some uses of lifting
can be simulated other ways within the system, though the limits of
this are still under investigation.  We also include a construct
$\delta$, for deriving a contradiction from a proof that
lambda-encoded true equals lambda-encoded false.  This also
compensates somewhat for the lack of lifting.

@@

In typing rules, when the type of an introduction form is checked or the type of an
elimination form is synthesized, we use call-by-name normalization, written
\(\leadstoc\) and \(\leadstocs\) for the reflexive transitive closure, to put
types in weak head normal form, revealing type constructors.
We abbreviate the conjunction ``term \(t\) synthesizes some type, and that type
call-by-name reduces to another type'' with the symbol \(\tpsynthleads\),
defined formally at the top of Figure~\ref{fig:tp}.
When a redex (reducible expression) occurs such that the argument to a type is a
term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
substitution with annotations: \([t_1/x]^{T_1}\ T_2\).

Call-by-name reduction is \emph{not} what underpins the
equivalence relation \(\cong\) for types and kinds, which is full
\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms),
both of which are modulo erasure of annotations in terms.
The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
the \emph{extraction} function for the Implicit Calculus of Constructions given
by~\cite{BB08_ICC-as-a-Programming-Language} adapted to CDLE.
To understand the role of erasure, recall that the type theory is \emph{extrinsic} (a.k.a.\
Curry-style), and hence we only consider erasures \(|t|\) of terms when testing
convertibility.
This is lifted to the conversion relation on types \(T \cong T'\) and kinds
\(\kappa \cong \kappa'\), whose rules are given in
Figure~\ref{fig:conv}.

@@

\begin{figure}
  \[
    \begin{array}{c}
      [t/x]^T\ =\ [\chi\ T\ \mhyph\ t/x]
      \\ \\ \\
  \begin{array}{cc}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T'}\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} \tpsynth \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
    \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

@@

@@

@@

>>>>

\newcommand{\choice}[0]{\zeta}
\newcommand{\elcap}[0]{\cap}
\newcommand{\abs}[4]{{#1}\, #2\! : \! #3.\, #4}
\newcommand{\absu}[3]{{#1}\, #2.\, #3}
\mathchardef\mhyph="2D % Define a "math hyphen"
\newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
\newcommand{\leadstoc}[0]{\ensuremath{\leadsto_{\mathbf{n}}}}
\newcommand{\leadstocs}[0]{\ensuremath{\leadsto_{\mathbf{n}}^*}}
\newcommand{\nleadstoc}[0]{\ensuremath{\nleadsto_{\mathbf{n}}}}
\newcommand{\tpcheck}[0]{\Leftarrow}
\newcommand{\tpsynth}[0]{\Rightarrow}
\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadstocs}{\Rightarrow}}}
\newcommand{\cbe}[0]{c\beta\eta}
\newcommand{\utp}[0]{\mathcal{U}}
\newcommand{\startcase}[1]{\vspace{#1} \noindent\textbf{\underline{Case:}}}
\newcommand{\startsubcase}[1]{\vspace{#1} \noindent\textbf{\underline{Subcase:}}}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{proposition}[theorem]{Proposition}

@@

\author{Aaron Stump, Christa Jenkins \\
  Computer Science \\
  The University of Iowa \\
  \texttt{aaron-stump@uiowa.edu},
  \texttt{christa-jenkins@uiowa.edu}}

@@

The type theory of Cedille is called the Calculus of Dependent Lambda
Eliminations (CDLE).  This document presents the version of CDLE as of
April 13, 2021.  We have made many changes from the first paper on CDLE~\cite{stump17},
mostly in the form of dropping constructs we discovered (to our
surprise) could be derived~\cite{stump18}. We have also omitted
\emph{lifting} -- a technique for large eliminations with lambda
encodings -- in this document's version of CDLE.  Some uses of lifting
can be simulated other ways within the system, though the limits of
this are still under investigation (see
\cite{JMS21_Simulating-Large-Elims-Cedille}).  We also include a construct 
$\delta$, for deriving a contradiction from a proof that
lambda-encoded true equals lambda-encoded false.  This also
compensates somewhat for the lack of lifting.

@@

In typing rules, when the type of an introduction form is checked or the type of an
elimination form is synthesized, we use call-by-name normalization, written
\(\leadstoc\) and \(\leadstocs\) for the reflexive transitive closure, to put
types in weak head normal form, revealing type constructors.
Note that with \(\leadsto\) we do not reduce under binders
(\(\lambda\),\(\Pi\),\(\forall\)) or reduce terms.
We abbreviate the conjunction ``term \(t\) synthesizes some type, and that type
call-by-name reduces to another type'' with the symbol \(\tpsynthleads\),
defined formally at the top of Figure~\ref{fig:tp}.
When a redex (reducible expression) occurs such that the argument to a type is a
term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
substitution with annotations: \(T_2[t_1/x]^{T_1}\).

Call-by-name reduction is \emph{not} what underpins the
equivalence relation \(\cong\) for types and kinds, which is full
\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms),
both of which are modulo erasure of annotations in terms.

The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
the \emph{extraction} function for the Implicit Calculus of Constructions given
by~\cite{BB08_ICC-as-a-Programming-Language} adapted to CDLE.
To understand the role of erasure, recall that the type theory is \emph{extrinsic} (a.k.a.\
Curry-style), and hence we only consider erasures \(|t|\) of terms when testing
convertibility.
This is lifted to the conversion relation on types \(T \cong T'\) and kinds
\(\kappa \cong \kappa'\), whose rules are given in
Figure~\ref{fig:conv}.

@@

\begin{figure}
  \[
    \begin{array}{c}
      [t/x]^T\ =\ [\chi\ T\ \mhyph\ t/x]
      \\ \\ \\
  \begin{array}{cc}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth \kappa[t/x]^{T'}}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth \kappa_1[T_2/X]}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} \tpsynth \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
    \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

@@

\begin{figure}
  \[
    \begin{array}{c}
      \fbox{\(T \rightsquigarrow T'\)}
      \quad \fbox{\(\kappa \rightsquigarrow \kappa'\)}
      \\ \\
      \begin{array}{cc}
        \infer{
         (\abs{\lambda}{x}{T_1}{T_2})\ t
         \rightsquigarrow
         T_2 [t/x]^{T_1}
        }{}
        &
          \infer{
           (\abs{\lambda}{X}{\kappa}{T})\ \cdot T'
           \rightsquigarrow
           T [T'/X]
          }{}
        \\ \\
        \infer{
         T_1\ t
         \rightsquigarrow
         T_1'\ t
        }{
         T_1 \rightsquigarrow T_1'
        }
        \\ \\
          \infer{
           T_1 \cdot T_2
           \rightsquigarrow
           T_1' \cdot T_2
          }{
           T_1 \rightsquigarrow T_1'
          }
        & 
        \infer{
         T_1 \cdot T_2
         \rightsquigarrow T_1 \cdot T_2'
        }{
         T_2 \rightsquigarrow T_2'
        }
        \\ \\
        \infer{
         \Gamma \vdash \abs{\lambda}{x}{T_1}{T_2} \leadsto \abs{\lambda}{x}{T_1'}{T_2}
        }{
         T_1 \leadsto T_1'
        }
        &
        \infer{
         \abs{\lambda}{x}{T_1}{T_2}
         \rightsquigarrow
         \abs{\lambda}{x}{T_1}{T_2'}
        }{
         T_2 \rightsquigarrow T_2'
        }
        \\ \\
        \infer{
         \abs{\lambda}{X}{\kappa}{T}
         \leadsto \abs{\lambda}{X}{\kappa'}{T}
        }{
         \kappa \leadsto \kappa'
        }
        &
        \infer{
         \abs{\lambda}{X}{\kappa}{T}
         \rightsquigarrow
         \abs{\lambda}{X}{\kappa}{T'}
        }{
         T \rightsquigarrow T'
        }
        \\ \\
          \infer{
           \abs{\mathcal{Q}}{x}{T_1}{T_2}
           \rightsquigarrow
           \abs{\mathcal{Q}}{x}{T_1'}{T_2}
          }{
           T_1 \rightsquigarrow T_1'
           \quad \mathcal{Q} \in \{\Pi,\forall,\iota\}
          }
        &
        \infer{
         \abs{\mathcal{Q}}{x}{T_1}{T_2}
         \rightsquigarrow
         \abs{\mathcal{Q}}{x}{T_1}{T_2'}
        }{
         T_2 \rightsquigarrow T_2'
         \quad \mathcal{Q} \in \{\Pi,\forall,\iota\}
        }
        \\ \\
          \infer{
           \abs{\forall}{X}{\kappa}{T}
           \rightsquigarrow
           \abs{\forall}{X}{\kappa'}{T}
          }{
           \kappa \rightsquigarrow \kappa'
          }
        & 
        \infer{
         \abs{\forall}{X}{\kappa}{T}
         \rightsquigarrow
         \abs{\forall}{X}{\kappa}{T'}
        }{
         T \rightsquigarrow T'
        }
        \\ \\ \\
        \infer{
         \abs{\Pi}{x}{T}{\kappa}
         \rightsquigarrow
         \abs{\Pi}{x}{T'}{\kappa}
        }{
         T \rightsquigarrow T'
        }
        &
          \infer{
           \abs{\Pi}{x}{T}{\kappa}
           \rightsquigarrow
           \abs{\Pi}{x}{T}{\kappa'}
          }{
           \kappa \rightsquigarrow \kappa'
          }
        \\
        \infer{
         \abs{\Pi}{X}{\kappa_1}{\kappa_2}
         \rightsquigarrow
         \abs{\Pi}{X}{\kappa_1'}{\kappa_2}
        }{
         \kappa_1 \rightsquigarrow \kappa_1'
        }
        &
          \infer{
           \abs{\Pi}{X}{\kappa_1}{\kappa_2}
           \rightsquigarrow
           \abs{\Pi}{X}{\kappa_1}{\kappa_2'}
          }{
           \kappa_2 \rightsquigarrow \kappa_2'
          }
      \end{array}
    \end{array}
  \]
  \caption{Term-free reduction of types and kinds}
  \label{fig:tp-reduce}
\end{figure}

@@

Lack of normalization for terms in general notwithstanding, it is still
important to confirm that the process of reducing types to weak head normal form
is itself terminating.
\begin{theorem}[Normalization of types under weak head normal form reduction]
  \label{thm:type-norm}
  \
  \begin{enumerate}
  \item If \(\Gamma \vdash \kappa\) then there exists \(\kappa'\) such that
    \(\kappa \leadsto^* \kappa' \nleadsto\)
  \end{enumerate} If \(\Gamma \vdash T : \kappa\) then there exists \(T'\) such
  that \(T \leadsto^* T' \nleadsto\).
\end{theorem}

@@

\section{Proof of Theorem~\ref{thm:type-norm}}
The proof proceeds by translation of the type language into a bespoke simply
typed lambda calculus which preserves, almost on-the-nose, the computation steps of
interest.
The typing rules are given in Figure~\ref{fig:bespoke-stlc}, which are to be
taken as implicitly defining the grammar, though note that term variables may be
both upper and lowercase.
The calculus has two base types: the unital \(\mathit{Top}\) with constructor
\(\mathit{top}\), and \(\mathit{\star}\) with constructor \(\mathit{con}\).
It also has two type constructors: the standard arrow \(\to\) for function types
and Cartesian products with \(\times\).
The reduction rules are standard, and so it is clear that this calculus enjoys
strong normalization and subject reduction.

\begin{figure}
  \[
    \begin{array}{ccc}
      \infer{
       \Gamma \vdash X : \tau
      }{
       (X\!:\!\tau) \in \Gamma
      }
      &
        \infer{
         \Gamma \vdash \mathit{top} : \mathit{Top}
        }{}
      &
        \infer{
         \Gamma \vdash \mathit{con}(x\!:\!\tau,t_1,t_2) : \star
        }{
         \Gamma \vdash t_1 : \tau_1
         \quad \Gamma,x\!:\!\tau \vdash t_2 : \tau_2
        }
      \\ \\
      \infer{
       \Gamma \vdash \abs{\lambda}{x}{\tau_1}{t} : \tau_1 \to \tau_2
      }{
       \Gamma, x\!:\!\tau_1 \vdash t : \tau_2
      }
      & &
        \infer{
         \Gamma \vdash t_1\ t_2 : \tau_2
        }{
         \Gamma \vdash t_1 : \tau_1 \to \tau_2
         \quad \Gamma \vdash t_2 : \tau_1
        }
      \\ \\
      \infer{
       \Gamma \vdash (t_1,t_2) : \tau_1 \times \tau_2
      }{
       \Gamma \vdash t_1 : \tau_1
       \quad \Gamma \vdash t_2 : \tau_2
      }
      & &
        \infer{
         \Gamma \vdash \mathit{proj}_i\ t : \tau_i
        }{
         i \in \{1,2\}
         \quad \Gamma \vdash t : \tau_1 \times \tau_2
        }
    \end{array}
  \]
  \caption{Bespoke STLC: typing}
  \label{fig:bespoke-stlc}
\end{figure}

The translation of CDLE's kind and type language to this calculus is shown in
Figure~\ref{fig:bespoke-trans}, and is inspired by the translation of CC to
System F\(_\omega\) by Geuvers and
Nederhof~\cite{GN91_Modular-Proof-of-Strong-Normalization-CC}.
Since reduction can also occur in kinds, we need two translations: one
into the a type of this calculus with \(\interp{-}_{\text{K}}\), and one into
the language of terms with \(\interp{-}\).
For the translation of CDLE kinds into types of the calculus, \(\Pi\)-kinds are
mapped to arrows, with quantification over terms translating to a domain of
\(\mathit{Top}\); for the translation of kinds into terms, we use the
constructor \(\mathit{con}\), translating the classifier of the bound variable
as both a type and a term of the calculus.

In types, type variables are preserved, lambda abstractions are translated to
lambda abstractions, where we introduce an additional redex to model reductions
in annotations on the bound variable and where again term parameters of any type
become parameters of type \(\mathit{Top}\), applications are translated to
applications where term arguments are replaced by the constant \(\mathit{top}\).
All other type constructs are translated to expressions involving
\(\mathit{con}\), where again the classifier ascribed to the quantified term (if
applicable) is translated as both a type and a term of the calculus.
Contexts are translated in the obvious way.

\begin{figure}
  \[
    \begin{array}{lcl}
      \interp{\star}_{\text{K}}
      & = 
      & \star
      \\ \interp{\abs{\Pi}{x}{T}{\kappa}}_{\text{K}}
      & =
      & \mathit{Top} \to \interp{\kappa}_{\text{K}}
      \\ \interp{\abs{\Pi}{X}{\kappa_1}{\kappa_2}}_{\text{K}}
      & =
      & \interp{\kappa_1}_{\text{K}} \to \interp{\kappa_2}_{\text{K}}
      \\ \\
      \interp{\star}
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\mathit{top},\mathit{top})
      \\ \interp{\abs{\Pi}{x}{T}{\kappa}}
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa})
      \\ \interp{\abs{\Pi}{X}{\kappa_1}{\kappa_2}}
      & =
      & \mathit{con}(X\!:\!\interp{\kappa_1}_{\text{K}}, \interp{\kappa_1},\interp{\kappa_2})
      \\ \\
      \interp{X}
      & =
      & X
      \\ \interp{\abs{\forall}{X}{\kappa}{T}}
      & =
      & \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
      \\ \interp{\abs{\forall}{x}{T_1}{T_2}}
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
      \\ \interp{\abs{\Pi}{x}{T_1}{T_2}}
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
      \\ \interp{\abs{\iota}{x}{T_1}{T_2}}
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
      \\ \interp{\{ t_1 \simeq t_2 \}}
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\mathit{top},\mathit{top})
      \\ \interp{\abs{\lambda}{x}{T_1}{T_2}}
      & =
      & \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2})}
      \\ \interp{\abs{\lambda}{X}{\kappa}{T}}
      & =
      & \abs{\lambda}{X}{\interp{\kappa}_{\text{K}}}{\mathit{proj_2}\ (\interp{\kappa},\interp{T})}
      \\ \interp{T\ t}
      & =
      & \interp{T}\ \mathit{top}
      \\ \interp{T_1 \cdot T_2}
      & =
      & \interp{T_1}\ \interp{T_2}
      \\ \\
      \interp{\cdot}
      & =
      & \cdot
      \\ \interp{\Gamma,X\!:\!\kappa}
      & =
      & \interp{\Gamma},X\!:\!\interp{\kappa}_{\text{K}}
      \\ \interp{\Gamma,x\!:\!T}
      & =
      & \interp{\Gamma},x\!:\!\mathit{Top}
    \end{array}
  \]
  \caption{Translation of kinds, types, and contexts}
  \label{fig:bespoke-trans}
\end{figure}

Straightaway, we see that, when translated as types, all CDLE kinds (well-formed
or not) are translated to well-formed types of this calculus, and
well-formedness of contexts is preserved by this translation.
It is also not hard to see that convertible kinds are translated to equal types,
and that substitution in kinds is a no-op.
\begin{proposition}
  \label{thm:stlc-kinds}
  \ \\
  \begin{enumerate}
  \item \(\interp{\kappa}_{\text{K}}\ \mathit{wf}\) for all \(\kappa\)

  \item If \(\vdash \Gamma\) then \(\mathit{DV}(\Gamma) =
    \mathit{DV}(\interp{\Gamma})\) and for all \((X\!:\!\kappa) \in \Gamma\),
    \((X\!:\!\interp{\kappa}) \in \interp{\Gamma}\)
  \item If \(\kappa_1 \cong \kappa_2\) then \(\interp{\kappa_1}_{\text{K}} = \interp{\kappa_2}_{\text{K}}\)
  \item \(\interp{\kappa[T/X]}_{\text{K}} = \interp{\kappa}_{\text{K}} = \interp{\kappa}_{\text{K}}
    [\interp{T}/X]\)
  \item \(\interp{\kappa[t/x]}_{\text{K}} = \interp{\kappa}_{\text{K}} = \interp{\kappa}_{\text{K}}[\mathit{top}/x]\)
  \end{enumerate}
\end{proposition}
\begin{proof}
  Each is provable by a simple inductive argument.
\end{proof}

We can also confirm that substitution is homomorphic for the translation.
\begin{lemma}
  \label{thm:stlc-subst-type}
  For all \(T\),\(T'\),\(\kappa\), and \(X\):
  \begin{enumerate}
  \item  \(\interp{T[T'/X]} = \interp{T}[\interp{T'}/X]\)

  \item \(\interp{\kappa[T'/X] = \interp{\kappa}[\interp{T'}/X]}\)
  \end{enumerate}

\end{lemma}
\begin{proof}
  By mutual induction. Proceed by cases of \(T\) and \(\kappa\).

  \startcase{.2cm} \(\kappa = \star\). Immediate.

  \startcase{.2cm} \(\kappa = \abs{\Pi}{x}{T}{\kappa'}\).
  Calculate
  \[
    \begin{array}{lclr}
      \interp{\kappa[T'/X]}
      & =
      & \interp{(\abs{\Pi}{x}{T}{\kappa'})[T'/X]}
      &
      \\
      & =
      & \interp{\abs{\Pi}{x}{T[T'/X]}{\kappa'[T'/X]}}
      & \text{def.\ subst}
      \\
      & =
      & \mathit{con}(x\!:\!\mathit{Top}, \interp{T[T'/X]},
        \interp{\kappa'[T'/X]})
      & \text{def.\ translation}
      \\
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T}[\interp{T'}/X],\interp{\kappa'}[\interp{T'}/X])
      & \text{IH}
      \\
      & =
      &
        \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa'})[\interp{T'}/X]
      & \text{def.\ subst}
      \\
      & =
      & \interp{\abs{\Pi}{x}{T}{\kappa'}}[\interp{T'}/X]
      & \text{def.\ translation}
      \\
      &
      &
      & \qed
    \end{array}
  \]

  \startcase{.2cm} \(\kappa = \abs{\Pi}{Y}{\kappa_1}{\kappa_2}\).
  Calculate
  \[
    \begin{array}{lclr}
      \interp{\kappa[T'/X]}
      & =
      & \interp{(\abs{\Pi}{Y}{\kappa_1}{\kappa_2})[T'/X]}
      &
      \\
      & =
      & \interp{\abs{\Pi}{Y}{\kappa_1[T'/X]}{\kappa_2[T'/X]}}
      & \text{def.\ subst}
      \\
      & =
      & \mathit{con}(Y\!:\!\interp{\kappa_1[T'/X]}_{\text{K}}, \interp{\kappa_1[T'/X]},
        \interp{\kappa_2[T'/X]})
      & \text{def.\ translation}
      \\
      & =
      & \mathit{con}(Y\!:\!\interp{\kappa_1}_{\text{K}}, \interp{\kappa_1[T'/X]},
        \interp{\kappa_2[T'/X]})
      & \text{Proposition~\ref{thm:stlc-kinds}(4)}        
      \\
      & =
      & \mathit{con}(Y\!:\!\interp{\kappa_1}_{\text{K}}, \interp{\kappa_1}[\interp{T'}/X],
        \interp{\kappa_2}[\interp{T'}/X])
      & \text{IH}
      \\
      & =
      &
        \mathit{con}(Y\!:\!\interp{\kappa_1}_{\text{K}},\interp{\kappa_1},\interp{\kappa_2})[\interp{T'}/X]
      & \text{def.\ subst}
      \\
      & =
      & \interp{\abs{\Pi}{Y}{\kappa_1}{\kappa_2}}[\interp{T'}/X]
      & \text{def.\ translation}
      \\
      &
      &
      & \qed
    \end{array}
  \]

  \startcase{.2cm} \(T = X\). Immediate.

  \startcase{.2cm} \(T = Y\) with \(Y \neq X\). Immediate.

  \startcase{.2cm} \(T = \abs{\forall}{Y}{\kappa}{T''}\).
  Calculate
  \[
    \begin{array}{lclr}
      \interp{T[T'/X]}
      & =
      & \interp{(\abs{\forall}{Y}{\kappa}{T''})[T'/X]}
      &
      \\
      & =
      & \interp{\abs{\forall}{Y}{\kappa[T'/X]}{T''[T'/X]}}
      & \text{def.\ subst}
      \\
      & =
      & \mathit{con}(Y\!:\!\interp{\kappa[T'/X]}_{\text{K}},\interp{\kappa[T'/X]},\interp{T''[T'/X]})
      & \text{def.\ translation}
      \\
      & =
      & \mathit{con}(Y\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa[T'/X]},\interp{T''[T'/X]})
      & \text{Proposition~\ref{thm:stlc-kinds}(4)}
      \\
      & =
      & \mathit{con}(Y\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa}[\interp{T'}/X],\interp{T''}[\interp{T'}/X])
      & \text{IH}
      \\
      & =
      &
        \mathit{con}(Y\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T''})[\interp{T'}/X]
      & \text{def.\ subst}
      \\
      & =
      & \interp{\abs{\forall}{Y}{\kappa}{T''}}[\interp{T'}/X]
      & \text{def.\ translation}
      \\
      &
      &
      & \qed
    \end{array}
  \]

  \startcase{.2cm} \(T = \abs{\forall}{x}{T_1}{T_2}\).
  Calculate
  \[
    \begin{array}{lclr}
      \interp{T[T'/X]}
      & =
      & \interp{(\abs{\forall}{x}{T_1}{T_2})[T'/X]}
      &
      \\
      & =
      & \interp{\abs{\forall}{x}{T_1[T'/X]}{T_2[T'/X]}}
      & \text{def.\ subst}
      \\
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1[T'/X]},\interp{T_2[T'/X]})
      & \text{def.\ translation}
      \\
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1}[\interp{T'}/X],\interp{T_2}[\interp{T'}/X])
      & \text{IH}
      \\
      & =
      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})[\interp{T'}/X]
      & \text{def.\ subst}
      \\
      & =
      & \interp{\abs{\forall}{x}{T_1}{T_2}}[\interp{T'}/X]
      & \text{def.\ translation}
      \\
      &
      &
      & \qed
    \end{array}
  \]

  \startcase{.2cm} \(T = \abs{\Pi}{x}{T_1}{T_2}\). Similar to the above case.

  \startcase{.2cm} \(T = \abs{\iota}{x}{T_1}{T_2}\). Similar to the above case.

  \startcase{.2cm} \(T = \{t_1 \simeq t_2\}\). Immediate.

  \startcase{.2cm} \(T = \abs{\lambda}{x}{T_1}{T_2}\).
  Calculate
  \[
    \begin{array}{lclr}
      \interp{T[T'/X]}
      & =
      & \interp{(\abs{\lambda}{x}{T_1}{T_2})[T'/X]}
      &
      \\
      & =
      & \interp{\abs{\lambda}{x}{T_1[T'/X]}{T_2[T'/X]}}
      & \text{def.\ subst}
      \\
      & =
      & \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1[T'/X]},\interp{T_2[T'/X]})}
      & \text{def.\ translation}
      \\
      & =
      & \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1}[\interp{T'}/X],\interp{T_2}[\interp{T'}/X])}
      & \text{IH}
      \\
      & =
      & \interp{T}[\interp{T'}/X]
      & \text{def.\ subst}
    \end{array}
  \]

  \startcase{.2cm} \(T = \abs{\lambda}{X}{\kappa}{T_2}\).
  Similar to the above case, appealing to Proposition~\ref{thm:stlc-kinds}(4)

  \startcase{.2cm} \(T = T_1\ t\), calculate
  \[
    \begin{array}{lclr}
      \interp{T[T'/X]}
      & = 
      & \interp{(T_1\ t)[T'/X]}
      &
      \\
      & =
      & \interp{T_1[T'/X]\ t[T'/X]}
      & \text{def.\ subst}
      \\
      & =
      & \interp{T_1[T'/X]}\ \mathit{top}
      & \text{def.\ translation}
      \\
      & =
      & \interp{T_1}[\interp{T'}/X]\ \mathit{top}
      & \text{IH}
      \\
      & 
      & \interp{T}[\interp{T'}/X]
      & \text{def.\ subst}
    \end{array}
  \]

  \startcase{.2cm} \(T = T_1 \cdot T_2\).
  Similar to the above case.
\end{proof}

\begin{lemma}
  \label{thm:stlc-subst-term}
  \
  \begin{enumerate}
  \item \(\interp{\kappa[t/x]} = \interp{\kappa}[\mathit{top}/x]\)
  \item \(\interp{T[t/x]} = \interp{T}[\mathit{top}/x]\) for all \(T\), \(t\), and \(x\).
  \end{enumerate}

\end{lemma}
\begin{proof}
  By inspection, the translation removes all occurrences of free term variables
  from \(T\) and \(\kappa\).
\end{proof}

The main effort left is to show the translation preserves the static and
term-skipping dynamic semantics of CDLE's type language.

\begin{lemma}
  \label{thm:stlc-static-sound}
  \ 
  \begin{enumerate}
  \item If \(\Gamma \vdash \kappa\) then \(\interp{\Gamma} \vdash
    \interp{\kappa} : \star\)
